---
title: "HRAttrition2"
output: html_document
date: "2024-11-15"
---

# Importing the Data
```{r pressure, echo=FALSE}
library(dplyr)

hr_df <- read.csv("WA_Fn-UseC_-HR-Employee-Attrition.csv")
glimpse(hr_df)


```

#Initial Data Exploration 
This section will perform data exploration of "raw" data set that has been imported.
But first, data type for some variables will be converted in order to make exploration easier.

## Changing 'character' columns to factor/numeric

```{r pressure, echo=FALSE}

hr_df$Attrition = as.factor(hr_df$Attrition )
hr_df$Department = as.factor(hr_df$Department )
hr_df$EducationField = as.factor(hr_df$EducationField)
hr_df$JobRole =  as.factor(hr_df$JobRole)
hr_df$Over18 =  as.factor(hr_df$Over18)
hr_df$OverTime =  as.factor(hr_df$OverTime)
hr_df$BusinessTravel =  as.factor(hr_df$BusinessTravel)
hr_df$MaritalStatus =  as.factor(hr_df$MaritalStatus)

glimpse(hr_df)

```

# Dataset Summary 

```{r pressure, echo=FALSE}
summary(hr_df)
```

#Data Cleaning

## Detecting null values 

```{r pressure, echo=FALSE}

colSums(is.na(hr_df))
```
# Analysis 

## Load Libraries 

```{r pressure, echo=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)

```

### Age and Attrition 

```{r pressure, echo=FALSE}
age_attr <- hr_df %>%
  group_by(Age, Attrition) %>%
  summarise(Counts = n()) %>%
  ungroup()

plot1 <-ggplot(age_attr, aes(x=Age, y= Counts, colour = Attrition )) + 
  geom_line() +
  labs(title = "Impact of Age on Employee Attrition Rate") + 
  theme_grey()

plot1

```

### Business Travel and Attrition 

```{r pressure, echo=FALSE}
btravel_attr <- hr_df %>%
  group_by(BusinessTravel, Attrition) %>%
  summarise(Counts = n()) %>%
  ungroup()

plot2 <-ggplot(btravel_attr, aes(x=BusinessTravel, y= Counts, fill = Attrition )) + 
  geom_bar(stat="identity", position = "dodge") +
  labs(title = "Impact of Business Travel on Employee Attrition Rate") + 
  theme_grey()

plot2

```
```{r pressure, echo=FALSE}
br_attr <- hr_df %>%
  group_by(BusinessTravel, Attrition) %>%
  summarise(Counts = n(), .groups = 'drop') %>%
  group_by(BusinessTravel) %>%
  mutate(Perc = round(Counts / sum(Counts) * 100, 2)) %>%
  ungroup()

ggplot(br_attr, aes(x = factor(BusinessTravel), y = Perc, fill = Attrition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Attrition Percentage by Business Travel",
       x = "Business Travel",
       y = "Attrition Percentage (%)",
       fill = "Attrition") +
  theme_minimal()

```

### Department and Attrition 

```{r pressure, echo=FALSE}
dept_attr <- hr_df %>%
  group_by(Department, Attrition) %>%
  summarise(Counts = n()) %>%
  ungroup()

plot4 <-ggplot(dept_attr, aes(x=Department, y= Counts, fill = Attrition )) + 
  geom_bar(stat="identity", position = "dodge") +
  labs(title = "Impact of Department on Employee Attrition Rate") + 
  theme_minimal()
plot4

```

```{r pressure, echo=FALSE}
sales_attr <-hr_df %>%
  group_by(Department, Attrition) %>%
  summarise(Counts=n()) %>%
  ungroup() %>%
  filter(Department == "Sales") %>%
  mutate(Perc = round(Counts / sum(Counts)*100, 2))

sales_attr

```

```{r pressure, echo=FALSE}

research_attr <-hr_df %>%
  group_by(Department, Attrition) %>%
  summarise(Counts=n()) %>%
  ungroup() %>%
  filter(Department == "Research & Development") %>%
  mutate(Perc = round(Counts / sum(Counts)*100, 2))

research_attr

bind_rows(sales_attr,research_attr)
```

```{r pressure, echo=FALSE}
dept_attr_pec1 <- hr_df %>%
  group_by(Department, Attrition) %>%
  summarise(Counts = n(), .groups = 'drop') %>%
  mutate(Perc = round(Counts / sum(Counts) * 100, 2)) %>%
  arrange(Department, desc(Counts))

dept_attr_pec1
```

```{r pressure, echo=FALSE}
dpt_attr1 <- hr_df %>%
  group_by(Department, Attrition) %>%
  summarise(Counts = n(), .groups = 'drop') %>%
  group_by(Department) %>%
  mutate(Perc = round(Counts / sum(Counts) * 100, 2)) %>%
  ungroup()


ggplot(dpt_attr1, aes(x = factor(Department), y = Perc, fill = Attrition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Attrition Percentage by Department",
       x = "Department",
       y = "Attrition Percentage (%)",
       fill = "Attrition") +
  theme_minimal()

```


### DistanceFromHome and Attrition 

```{r pressure, echo=FALSE}
travel_attr <- hr_df %>%
  group_by(DistanceFromHome, Attrition) %>%
  summarise(Counts = n()) %>%
  ungroup()

plot5 <-ggplot(travel_attr, aes(x=DistanceFromHome, y= Counts, color = Attrition )) + 
  geom_line() +
  labs(title = "Impact of Distance From Home on Employee Attrition Rate") + 
  theme_minimal()
plot5

```

### EnvironmentSatisfaction and Attrition 

```{r pressure, echo=FALSE}
satis_attr <- hr_df %>%
  group_by(EnvironmentSatisfaction, Attrition) %>%
  summarise(Counts = n()) %>%
  ungroup()

plot6 <-ggplot(satis_attr, aes(x=EnvironmentSatisfaction, y= Counts, fill = Attrition )) + 
  geom_bar(stat = "identity") +
  labs(title = "Impact of Environment Satisfaction on Employee Attrition Rate") + 
  theme_minimal()
plot6

```

### JobSatisfaction and Attrition 

```{r pressure, echo=FALSE}

job_satisfaction_attr <- hr_df %>%
  group_by(JobSatisfaction, Attrition) %>%
  summarise(Counts = n(), .groups = 'drop') %>%
  group_by(JobSatisfaction) %>%
  mutate(Perc = round(Counts / sum(Counts) * 100, 2)) %>%
  ungroup()


# Create a bar chart for attrition percentages by job satisfaction level
ggplot(job_satisfaction_attr, aes(x = factor(JobSatisfaction), y = Perc, fill = Attrition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Attrition Percentage by Job Satisfaction Level",
       x = "Job Satisfaction Level",
       y = "Attrition Percentage (%)",
       fill = "Attrition") +
  theme_minimal()

```


### NumCompaniesWorked and Attrition 

```{r pressure, echo=FALSE}
num_attr <- hr_df %>%
  group_by(NumCompaniesWorked, Attrition) %>%
  summarise(Counts = n()) %>%
  ungroup()

plot8 <-ggplot(num_attr, aes(x=NumCompaniesWorked, y= Counts, fill = Attrition )) + 
  geom_area(position = "stack", alpha=0.6) +
  labs(title = "Number of Companies Worked on Employee Attrition Rate") + 
  theme_minimal()
plot8

```


### WorkLifeBalance and Attrition 


```{r pressure, echo=FALSE}

worklife_attr <- hr_df %>%
  group_by(WorkLifeBalance, Attrition) %>%
  summarise(Counts = n(), .groups = 'drop') %>%
  group_by(WorkLifeBalance) %>%
  mutate(Perc = round(Counts / sum(Counts) * 100, 2)) %>%
  ungroup()


ggplot(worklife_attr, aes(x = factor(WorkLifeBalance), y = Perc, fill = Attrition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Attrition Percentage by Work Life Balance",
       x = "Work Life Balance",
       y = "Attrition Percentage (%)",
       fill = "Attrition") +
  theme_minimal()

```

### Years At Company and Attrition 

```{r pressure, echo=FALSE}
year_attr <- hr_df %>%
  group_by(YearsAtCompany, Attrition) %>%
  summarise(Counts = n()) %>%
  ungroup()

plot10 <-ggplot(year_attr, aes(x=YearsAtCompany, y= Counts, color = Attrition )) + 
  geom_line() +
  labs(title = "Years at Company on Employee Attrition Rate") + 
  theme_minimal()

plot10

```


# Prediction 

## COrrelation HEAT MAP ## 

```{r pressure, echo=FALSE}
library(dplyr)
library(corrplot)
```
# Select relevant columns for X and Y 

```{r pressure, echo=FALSE}
X <- hr_df %>%
  select(-Attrition, -StandardHours, -EmployeeCount, - EmployeeNumber, -HourlyRate, -MonthlyRate) %>% select(where(is.numeric))
 
X
```


```{r pressure, echo=FALSE}
Y <- hr_df$Attrition

Y
```

```{r pressure, echo=FALSE}
cor_matrix<- cor(X, use ="complete.obs")
```


```{r pressure, echo=FALSE}
corrplot(cor_matrix, method = "color", col = colorRampPalette(c("yellow","white","green"))(200), tl.cex=0.7, tl.col = "black", number.cex = 0.5, is.corr=TRUE,  addCoef.col = "black", number.digits=2)

```

##Logistic Regression ##


```{r pressure, echo=FALSE}
#load library
library(caret)
library(glmnet)

#select relevant column for x and y
x<- hr_df %>%
  select(-Attrition, -StandardHours, -EmployeeCount)
y <- hr_df$Attrition


# Split the data into training and test sets

set.seed(0)  
trin_index <- createDataPartition(y, p= 0.75, list = FALSE)
x_train <-x[train_index, ]
x_test <- x[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]


# Train logistic regression mode

log_reg <-glmnet(as.matrix(x_train), as.factor(y_train), family = "binomial", lambda = 0.001, alpha =0)

# Predict and calculate training and test accuracy

y_train_pred <- predict(log_reg, newx = as.matrix(x_train), type = "class")
y_test_pred <- predict(log_reg,newx = as.matrix(x_test), type = "class")

train_accuracy <- mean(y_train_pred == y_train)
test_accuracy <- mean(y_test_pred == y_test)

# Output the results
cat('--------------------------------------------------------------------------\n')
cat('Logistic Regression:\n')
cat(sprintf('Training Model accuracy score: %.3f\n', train_accuracy))
cat(sprintf('Test Model accuracy score: %.3f\n', test_accuracy))
cat('--------------------------------------------------------------------------\n')
```

#Training Model

```{r pressure, echo=FALSE}
# Load necessary libraries
library(dplyr)
library(caret)
library(glmnet)
library(fastDummies)

str(hr_df)

# One-hot encode categorical variables in the dataset
data_dummies <- dummy_cols(df,
                           select_columns=c("BusinessTravel", "Department",
                                            "EducationField", "Gender",
                                            "JobRole", "MaritalStatus",
                                            "Over18", "OverTime", "Attrition"),
          remove_first_dummy = TRUE,  
        remove_selected_columns = TRUE) 


summary(data_dummies)

#fast dummies create blanks in the col names. remove blanks using make.names 
library(tidyverse)
colnames(data_dummies)

colnames(data_dummies) <- make.names(colnames(data_dummies))

# Select relevant features for X
features <- data_dummies %>%
  select( Age,DistanceFromHome,Education,EmployeeCount,EmployeeNumber,EnvironmentSatisfaction,JobInvolvement,JobLevel,JobSatisfaction,MonthlyIncome,NumCompaniesWorked,PercentSalaryHike,PerformanceRating
,RelationshipSatisfaction,StockOptionLevel,TotalWorkingYears,TrainingTimesLastYear,WorkLifeBalance,YearsAtCompany,YearsInCurrentRole,YearsSinceLastPromotion,YearsWithCurrManager,BusinessTravel_Travel_Frequently,BusinessTravel_Travel_Rarely,Department_Research...Development,Department_Sales,EducationField_Life.Sciences,EducationField_Marketing,EducationField_Medical,EducationField_Other,EducationField_Technical.Degree,Gender_Male,JobRole_Human.Resources,JobRole_Laboratory.Technician,JobRole_Manager,JobRole_Manufacturing.Director,JobRole_Research.Director,JobRole_Research.Scientist,JobRole_Sales.Executive,JobRole_Sales.Representative,MaritalStatus_Married,MaritalStatus_Single,Over18_,OverTime_Yes

)

# Convert features to matrix for logistic regression
x1 <- as.matrix(features)

# Define target variable y
y1 <- as.factor(data_dummies$Attrition_Yes)

# Split data into training and test sets
set.seed(0)  # For reproducibility
train_index <- createDataPartition(y, p = 0.75, list = FALSE)
x1_train <- x1[train_index, ]
x1_test <- x1[-train_index, ]
y1_train <- y1[train_index]
y1_test <- y1[-train_index]

# Predict and calculate accuracy on the trained model

log_reg1 <-glmnet(x1_train, y1_train, family = "binomial", lambda = 0.001, alpha = 0, maxit = 10000)

y1_train_pred <- predict(log_reg1, newx = x1_train, type = "class")
y1_test_pred <-predict(log_reg1, newx = x1_test, type = "class")

train_accuracy1 <- mean(y1_train_pred == y1_train)
test_accuracy1 <- mean(y1_test_pred == y1_test)


# Output the results
cat('--------------------------------------------------------------------------\n')
cat('Logistic Regression:\n')
cat(sprintf('Training Model accuracy score: %.3f\n', train_accuracy1))
cat(sprintf('Test Model accuracy score: %.3f\n', test_accuracy1))
cat('--------------------------------------------------------------------------\n')

```


```{r pressure, echo=FALSE}
# Extract coefficients for the trained model
coefficients <- as.matrix(coef(log_reg1, s = 0.001))

coefficients 

# Convert coefficients to a data frame for easy viewing
coefficients_df <- data.frame(
  Feature = rownames(coefficients),
  Coefficient = coefficients[, 1]
)

# Sort by absolute value of the coefficient 
coefficients_df <- coefficients_df %>%
  arrange(desc(abs(Coefficient)))

# Display the top impacting features
head(coefficients_df, 10)
```